
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Introduction

```{r, echo=FALSE, fig.alt='Data Management and Sharing, written by Carrie Wright and Candace Savonen', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.gd422c5de97_0_0")
```

```{r, echo=FALSE, fig.alt='This overall learning objectives of this course are to demonstrate how to: Understand what data sharing is and why is it important?, Effectively manage your data including the associated skills relating to data heavy projects, Maintain and write effective documentation, records, and metadata, Organize your project so that it is reproducible and well understood by others, Create a data management sharing plan consistent with NIH requirements, Identify and use computing resources that best fit your project needs, Store and submit your data to repositories', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.gd422c5de97_0_10")
```


## Target Audience  

The course is intended for individuals in in the cancer research community who want to learn the best practices and techniques for data management and sharing.

_This course is written for individuals who:_   

- Have or plan to have biomedical data they need to manage
- Have or plan to apply for NIH funding
- Work directly with the cancer research data or mentor those who work with data
- Have not had much training or background in data handling practices


```{r, echo=FALSE, fig.alt='For individuals who:Have or plan to have biomedical data they need to manage, Have or plan to have or apply for NIH funding, Either work directly with the data or help mentor those who work with data,Have not had much training or background in data handling practices', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.g1173f7473f7_0_0")
```

## Topics covered

This course covers how to properly manage and share data including:


```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.gfd7c7a0ba8_0_0")
```

- Understand what data sharing is and why is it important?
- Effectively manage your data including the associated skills relating to data heavy projects
- Maintain data privacy and comply with data privacy laws
- Maintain and write effective documentation
- Keep effective records that will help you track your project properly but securely
- Create good metadata that can enhance the use your data
- Organize your project so that it is reproducible and well understood by others
- Create a data management sharing plan consistent with NIH requirements
- Identify and use computing resources that best fit your project needs
- Store and submit your data to repositories


## Motivation

The cancer research discipline has evolved into an increasingly complex mix of datasets - research projects are typically cross-disciplinary and contain many types of data in various formats. They often involve multiple collaborators generating data across different sites, with different data standards and infrastructure used to generate the data. Therefore, it is more important than ever to be well-versed in the best practices of data management and sharing.

Not only is proper data management and sharing a necessity for cancer research projects to succeed in positively impacting cancer care, but it is now also increasingly a necessity to obtain funding as the NIH and other cancer research funders have implemented mandates that require you to proactively plan to manage and share your data.

As a member of the cancer research community, it is imperative that you maintain well-documented metadata and properly share your data. This will benefit you, your colleagues, and the larger community by broadening the reach of your data, enabling data reuse by others, and ultimately accelerating the pace of scientific discovery. This course aims to serve as a starting point to cover the basics of good data management and sharing practices.


## Curriculum  

**How to use this course:**
This course contains high-level concepts for data management and sharing and can be used as a reference of suggested best practices and associated skills needed for data management and sharing in biomedical research.

**Keep in mind:**
Scientific data and research projects come in many different forms, and some content in this course may not apply, especially as the research landscape evolves to adapt and support new technology, methods, and techniques. Therefore, the goal of this course is not to prescribe rigid rules for how to conduct research, but rather serve as a guide to approach data management and sharing in the spirit of the FAIR principles (Findable, Accessible, Interoperable, Reusable). We encourage you to continue to consult with data management experts to suit the needs of your particular project and/or research goals.

<div class = disclaimer>
`r config::get("disclaimer")`
</div>
