
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Introduction

```{r, echo=FALSE, fig.alt='Data Management and Sharing, written by Carrie Wright and Candace Savonen', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.gd422c5de97_0_0")
```

```{r, echo=FALSE, fig.alt='This overall learning objectives of this course are to demonstrate how to: Understand what data sharing is and why is it important?, Effectively manage your data including the associated skills relating to data heavy projects, Maintain and write effective documentation, records, and metadata, Organize your project so that it is reproducible and well understood by others, Create a data management sharing plan consistent with NIH requirements, Identify and use computing resources that best fit your project needs, Store and submit your data to repositories', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.gd422c5de97_0_10")
```


## Target Audience  

The course is intended for individuals in biomedical scientists and program managers who want to learn the best practices and techniques for data management and sharing.

_This course is written for individuals who:_   

- Have or plan to have biomedical data they need to manage
- Have or plan to have or apply for NIH funding
- Either work directly with the data or help mentor those who work with data
- Have not had much training or background in data handling practices


```{r, echo=FALSE, fig.alt='For individuals who:Have or plan to have biomedical data they need to manage, Have or plan to have or apply for NIH funding, Either work directly with the data or help mentor those who work with data,Have not had much training or background in data handling practices', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.g1173f7473f7_0_0")
```

## Topics covered

This course covers how to properly manage and share data including:


```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.gfd7c7a0ba8_0_0")
```

- Understand what data sharing is and why is it important?
- Effectively manage your data including the associated skills relating to data heavy projects
- Maintain and write effective documentation
- Keep effective records that will help you track your project properly but securely
- Create good metadata that can enhance the use your data
- Organize your project so that it is reproducible and well understood by others
- Create a data management sharing plan consistent with NIH requirements
- Identify and use computing resources that best fit your project needs
- Store and submit your data to repositories


```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit#slide=id.g1173f7473f7_0_0")
```

## Motivation

Biomedical projects are increasingly data based. Not only is proper data sharing and management a non negotiable necessity for data heavy projects but NIH and other grant funding institutions have mandates that require you to do so. But it can be difficult to be brought up to speed on these effective management especially as data requirements and best practices continue to change rapidly.

Effective data management and sharing can mean the difference between more funding or not. Well documented and properly shared data is not only the right thing to do as a part of a scientific community but it will help shed a spotlight on your work in the community. It also is not only helpful for yourself but for you and your future colleagues should you need to return to your work or build upon it (which is what science usually aims to do).

## Curriculum  

**How to use this course:**
This course contains high-level concepts for data management and sharing and can be used as a reference of suggested best practices and associated skills needed for data management and sharing in biomedical research.

**Keep in mind:**
Scientific data and research projects come in many different forms, and some content in this course may not apply, especially as the research landscape evolves to adapt and support new technology, methods, and techniques. Therefore, the goal of this course is not to prescribe rigid rules for how to conduct research, but rather serve as a guide to approach data management and sharing in the spirit of the FAIR principles (Findable, Accessible, Interoperable, Reusable). We encourage you to continue to consult with data management experts to suit the needs of your particular project and/or research goals.

<div class = disclaimer>
`r config::get("disclaimer")`
</div>
