# (PART\*) What is Data Sharing? {-}

In this section we will cover the following learning objectives:


```{r, fig.align='center', echo = FALSE, fig.alt= "Learning objectives: 1. Explain how data sharing encourages scientific advancement, 2. Describe the benefits of data sharing ", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit?slide=id.g2df9e1c7fe6_0_0#slide=id.g2df9e1c7fe6_0_0")
```

# Data Sharing is Important

Sharing data is critical for optimizing the advancement of scientific understanding. Now that labs all over the world are producing massive amounts of data, there are many discoveries that can be made by simply re-using/re-analyzing this existing data.

The concept of data re-use is so important, that in January 2023, after years of development, the NIH released new guidance and began requiring new practices for data management and sharing. See [the announcement here](https://grants.nih.gov/grants/guide/notice-files/NOT-OD-21-013.html).


See this [course about the NIH policy](https://hutchdatascience.org/NIH_Data_Sharing/) for more information about how to comply.

<div class = "warning">
Note that many institutes and funding agencies or mechanisms have requirements about how your data can be shared. Typically, data sharing of protected data also requires Institutional Review Board (IRB) approval before the study is conducted. Ensure that you are following those requirements before you share your data. A later section in this course will cover [data privacy](https://hutchdatascience.org/Proactive_Data_Management_and_Sharing/data-privacy.html).
</div>

There's so many excellent reasons to put your data in a repository, whether or not a journal requires it:  

**Sharing your data...**  

1. Makes your project more transparent and thus more likely to be trusted and cited. In fact, one study found that articles with links to the data used (in a repository) were cited more than articles without such information or other forms of data sharing [@colavizza_citation_2020].

```{r, fig.align='center', echo = FALSE, fig.alt= "Another researcher is downloading the data from a repository and says ‘These insights are so exciting! I can’t wait to look into this data even more!’", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g117c57cc481_0_636")
```

2. Helps reduce your own workload so your email inbox isn't overloaded by requests you probably don't have time to respond to.

```{r, fig.align='center', echo = FALSE, fig.alt= "Ruby is reading a journal article with data she is interested in. The journal article says ‘Data are available upon request by email’. Ruby sends an email that says ‘ The email is going to an inbox with 999,999,565473 emails in it and it is labeled ‘the corresponding author’s inbox’.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit?slide=id.g35294cebb05_0_0#slide=id.g35294cebb05_0_0")
```

3. Allows others to gain even more insights from your data which shows funders that your data will be used to its maximum potential.

```{r, fig.align='center', echo = FALSE, fig.alt= "Ruby has uploaded her data to a repository and now its being used by many other researchers. Ruby says to her funders, represented as a bank, ‘The data you funded is getting so much mileage!’", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g117c57cc481_1_37")
```

In summary, data sharing is the practice of making all appropriate data sources associated with a project available through systematic means. Broad data sharing is a critical step toward mitigating the reproducibility and replicability crisis. Not only that, but data sharing is economical, equitable, and frugal. New exciting findings can result from shared and reused data. Additionally, institutions and individuals with less resource availability can benefit from access to shared data.
