# Record keeping practices

In this section we will cover the following learning objectives:

```{r, fig.align='center', echo = FALSE, fig.alt= "This chapter will demonstrate how to: Identify methods for keeping records of communication, Explain how notebooks can be especially helpful for collaboration and record keeping."}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit?slide=id.g3896feb580f_34_1#slide=id.g3896feb580f_34_1")
```

Once you have your project rolling, it is important to keep good records of your work, your collaborators work, and your communication. Keeping good records takes time and discipline but it can save you more time and heartache in the end. Here are some suggestions for how to optimize your record keeping.



### **Keep organized records of work**

Record and communicate notes about your data collection and analyses. Be mindful of overwhelming your coworkers, but generally speaking provide extra information where possible. The more people that are aware of details about what samples were in what batch, the more likely important details are not missed or forgotten. For example, if you are sending data to a collaborator, send as much information as possible about how it was generated in the email in which you send it to them, even if you have already discussed the data. This can help ensure that no important details fall through the cracks. The best way we think you can do this in general is to use reports - one of our next suggestions.


```{r, fig.alt="Francis-the frustrated informaticist receives an email from Charlie that says: Hi Francis, attached are the data. Looking forward to the results next week! Francis thinks to herself: I have no idea what samples are in this data. She looks distressed.", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OU5qeRgN_fojGbcyu2qEdwlcKpDO6BveWtYW_u1Hqd4/edit#slide=id.gcee686a31e_0_596")
```

### **Keep organized records of communication**

Besides recording your work, keep a record of your communications. At a minimum organize your emails for projects into separate folders with easily recognizable titles to save yourself hassle later when something comes into question. However, we **highly** recommend that in addition for even better record keeping, you can use a [note-taking system](https://zapier.com/blog/best-note-taking-apps/). This could be as simple as a shared Google doc, or you could consider an app like [these](https://collegeinfogeek.com/best-note-taking-apps/) that are designed for note-taking.  With many of these you can also share your notes with research teammates and you can include report documents directly in your notes. Which brings us to our next point about using reports!


```{r, fig.alt=" Charlie, the unclear leader is looking through his email and sees identical emails that say: Hi Francis, attached are the data, on January 19th, March 15th, and April 16th. Charlie thinks to himself: What data did I send to Francis? Uh oh, I don't know. Thus Francis isn't the only one confused about what data Charlie sent.", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OU5qeRgN_fojGbcyu2qEdwlcKpDO6BveWtYW_u1Hqd4/edit#slide=id.gcee686a31e_0_622")
```

### **Use reports**

Instead of sending informal short emails (which are useful at some points in a workflow), we suggest intermittently sending lab reports with as much information about what was done and why. For informatics related work in R or Python (or other supported languages) we **highly** suggest using a method like [R markdown](https://rmarkdown.rstudio.com/lesson-10.html) or [Jupyter notebooks](https://jupyter.org/) to track what informatics steps you have performed and why. Beginning these reports with a short description of what raw data you used and when you received it can be critical for ensuring that you are using the correct data! We recommend this chapter about [reproducible reports using these tools](https://hutchdatascience.org/Tools_for_Reproducible_Workflows_in_R/reproducible-reports.html) if you are interested in more informaiton. 

It is also important that the experimental biologists make similar reports defining what reagents they used, when they performed the study, what samples were used, who performed the experiment, and any notes about unusual events, such as the electricity went out during the experiment, left the samples overnight but usually leave two hours, mouse #blank unexpectedly died so we lost this sample thus it is not included, or the dye seemed unusually faint in this gel.  


```{r, fig.alt="Hilda, the happy informaticist is reading a report sent from Harry that says: The data is from the samples sent tus on October 10th created using the Illumina Kit and process with... and it trails off. Hilda thinks to herself: Ah, Harry's reports ave me so much time! I think I'll have my dissertation done early! I think we will all be able to take a vacation!", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OU5qeRgN_fojGbcyu2qEdwlcKpDO6BveWtYW_u1Hqd4/edit#slide=id.gcf28d80132_0_305")
```



```{r, fig.alt="Charlie, the unclear leader and Jack, the lonely and only informaticist in his lab are having a meeting. Charlie says: I'm sorry Jack, we will need to redo the analysis with different data. It appears there was a miscommunication with our collaborator. Jack thinks to himself: Oh dear, It's looking like grad school will take 30 years. You can see dark clouds forming above them.", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OU5qeRgN_fojGbcyu2qEdwlcKpDO6BveWtYW_u1Hqd4/edit#slide=id.gcf28d80132_0_360")
```

In summary, we recommend the following record keeping tips:


```{r, fig.alt=" Record Keeping Tips: 1) Keep organized records of work - Take your own detailed notes - we suggest reports using R markdown or Jupyter notebooks 2) Keep organized records of communication - Keep emails organized in folders and name emails in a useful way to find later  - Consider a note-taking app to store emails, notes, and reports in one searchable place 3) Use reports and include as much detail as possible. Send these reports regularly to research teammates", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OU5qeRgN_fojGbcyu2qEdwlcKpDO6BveWtYW_u1Hqd4/edit#slide=id.gcd0b5dd2d8_1_40")
```
