# (PART\*) What is Data Sharing and Why is it Important? {-}

TODO: Summarize this and fix headers

# Computing Infrastructure

# Computing Resources

In this chapter we will describe the basics about data size and computing capacity. We will discuss the computing and storage requirements for many types of cancer related data, as well as options to perform informatics work that might require more intensive computing capacity than your personal computer.


```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1. Name units of size for binary data, 2. State the computing and storage capacity of typical computers today, 3. Determine the range of storage and computing capacity required for various bioinformatics studies, 4. Recognize different methods for performing intensive computations or storing large data, 5. Explain how server and cloud computing works", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf96b1d997a_0_9")
```


## Data Sizes

Recall that the smallest unit of data is a bit which is either a zero or a one. A group of 8 bits is called a byte, and most computers and phones, and software programs are constructed or designed in a way to accommodate groups of bytes at a time. For example a 32-bit machine can work with 4 bytes at a time and a 64-bit can work with 8 bytes at a time. But how big is a file that is 2 GB? When we sequence a genome, how large is that in terms of binary data? Can our local computer work with the size of data that we would like to work with?


First let's take a look at how the size of binary data is typically described and what this actually means in terms of bits and bytes:


```{r, fig.align='center', echo = FALSE, fig.alt= "Table of different binary data units showing the name, abbreviation, and size in bits or bytes, for example a Byte is abbreviated as B and this represents 8 bits, while Gigabyte is abbreviated GB and represents roughly 1 billion bytes", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_8")
```


```{r, fig.align='center', echo = FALSE, fig.alt= "Cartoon - One character says:Hey I'm so stoked. I have a super cool computer now, it has yodabytes. The other character says: Do you mean yottabytes. The original character says: Nope, I have lots of files about Yoda! The other character says: Oh dear...", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb6624a387_0_1")
```

Now that we know how to describe binary data sizes, let's next think about how much computing capacity typical computers have today.

## Computing Capacity

We have discussed a bit about CPUs and how they can help us perform more than one task at a time, but how many tasks can the CPU of an average computer do simultaneously these days? How much memory and storage do they typically have? What size of files can a typical computer handle? This information is sometimes called the **specs** of a computer.

These values will probably change very soon, and different computers vary widely, but currently:

- **Laptops** can often perform 4-8 CPU tasks at once, and typically range from 4-16 GB in memory and 250 GB-1 TB of storage.

This means that typical laptops can multitask quite well, have in some cases 16 gigabytes for random access memory to allow the CPU to work on relatively large tasks (as we can see from the previous table that GB are actually pretty large when you think about it), and possibly 1TB for the hard drive (and or SSD), meaning that you can store thousands of photos and files like PDFs, word documents etc. It turns out that you can store around 30,000 average size photos with 250GB, so a 1TB laptop can store quite a bit of data. Therefore overall, typical laptops today are pretty powerful compared to previous computers. Note that some programs require 16 or even 32 GB of memory to run.

- **Desktops** can perform and store similarly and sometimes to a degree better than a laptop for a similar price.  Since less work needs to be done to make the desktop small and portable, sometimes you can get better storage and performance for the same price as a laptop. However, desktops often have better graphics processing capacity and displays and that might make up for the price difference [@antonio_villas-boas_laptops_2019].  This might be important to consider if you are going to need to visually inspect many images. Another benefit is that you can also sometimes find desktops with larger memory and storage options right off the shelf than typical laptops. It is also generally easier to add more memory to a desktop than it is to add to a laptop [@antonio_villas-boas_laptops_2019]. However of course, desktops certainly aren't super portable!


* Some **phones** can compete with laptops by performing 6 CPU tasks at once and storing 6 GB in memory and 250 GB of storage.  


```{r, fig.align='center', echo = FALSE, fig.alt= "Cartoon - One character says: Hey, are you moving?. The other character says: Na, I’m just bringing my computer to work, I do it everyday. It shows the character with a wheel barrel with a desktop computer in it.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.g1084414e81e_0_0")
```



Check out this [link](https://www.apple.com/mac/compare/?modelList=iMac,MacBook-Pro-14,MacBook-Pro-16-2021) to compare the prices of different macs and this [link](https://www.hp.com/us-en/shop/slp/weekly-deals) to compare specs for PC computers from HP.

If you want to get really in-depth comparisons for PC or windows machines, check out this [link](https://www.userbenchmark.com/PCBuilder/Custom/S0-M1487712vsS0-M?tab=RAM) [@userbenchmark].



### Checking your computer capacity - Mac

So what about your computer? How do you know how many cores it has or how much memory and storage it has?

If you have a Mac, you can click on the apple symbol on the far left of your screen. Then click on the "About This Mac" button.

You might see something like this:

```{r, fig.align='center', echo = FALSE, fig.alt= "Result of pressing `About this Mac` button of a typical Mac today", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_69")
```

First we see the operating system is called Mojave.
Next we see that the processor (which we now know is the CPU) is a 2.6 GigaHertz (GHz) Intel Core i7 chip. This means that the processor or CPU can process 2,600,000,000 operations in a second (this is called a [clock cycle](http://www.techopedia.com/definition/5498/clock-cycle)) [@clock_cycle]. That's a lot compared to older computers which had clock cycle rate or [clock rate](https://en.wikipedia.org/wiki/Clock_rate) in the MegaHertz range in the 1980s [@clock_rate]!
If we look up more about this chip we would learn that it has 4 cores and has hyper-threading, which allows it to effectively perform 8 tasks at once [@hyperthreading].
Next we see that there is 16 Gigabytes of memory - this is how much RAM it has and also 2133 MegaHertz (aka 2.133 GHz) of low power double data rate random access memory (LPDDR3), this means that the RAM can process  2,133,000,000 commands every second [@RAM_speed; @mukherjee_ram_2019]. If you are interested you can checkout more about what this means at this blog post @scott_thornton_RAM. However, generally the amount of RAM is more important for assessing performance [@RAM_speed; @mukherjee_ram_2019].

If we click on the storage button at the top, we can learn about how much storage is available on the computer. If you hover over a section, it tells you what file are accounting for that section of storage that is already being used.

```{r, fig.align='center', echo = FALSE, fig.alt= "Mac storage information showing 1 TB capactity", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_80")
```

### Checking your computer capacity - Windows/PC


If you have a PC or Windows computer, the steps may vary depending on your operating system, but try the following:

```{r, fig.align='center', echo = FALSE, fig.alt= "Windows/PC Start button", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_96")
```

1. click the "Start" button - which looks like 4 squares together
2. click "Settings" button (gear-shaped)
3. click "System"
4. click "About"

See this [link](https://www.businessinsider.com/how-to-find-computer-specs-windows-10) for more information.


```{r, fig.align='center', echo = FALSE, fig.alt= "Windows/PC about section", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_90")
```

```{r, fig.align='center', echo = FALSE, fig.alt= " Example of a Windows/PC computer processor and memory information", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_107
")
```

Here we can see that this computer has an Intel(R) Core(TM) i7-4790K CPU @ 4.00 GHz 4.00 GHz chip and 16 Gigabytes of RAM.
If we look up this chip we can see that it has 4 cores and 8 threads (due to hyper-threading) allowing for 8 tasks at a time.

To find out more information about your storage click the "Storage" button within the "System" tab.

```{r, fig.align='center', echo = FALSE, fig.alt= "Windows/PC storage information showing 1 TB capactity", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_114
")
```

Here we can see that this computer has 466 GB + 465 GB = 932 GB across the two drives. The C drive is typically for the operating system, and the D drive is typically where you would install application programs and save files. There are 1000 GB in a TB, thus, this computer has about the same storage as the Mac that we just looked at.

## File Sizes

Now let's think about the files that we might need for our research, how big are files typically for genomic, imaging, and clinical research?

Recall this table from earlier about digital data size units:

```{r, fig.align='center', echo = FALSE, fig.alt= "Table of different binary data units showing the name, abbreviation, and size in bits or bytes, for example a Byte is abbreviated as B and this represents 8 bits, while Gigabyte is abbreviated GB and represents roughly 1 billion bytes", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_8")
```

### Genomic data file sizes

Genomic data files can be quite large and can require quite a bit of storage and processing power.

Here is an image of sizes of some common file types:

```{r, fig.align='center', echo = FALSE, fig.alt= "Table of file types for genomics data, whole genome sequencing can become larger than the capacity of your computer with less than 20 samples! Even whole exome sequenceing can already require more than 44% of a 1TB hard drive for just 20 samples. Note that these are approximate values.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_0")
```



### Imaging Data File Sizes

Imaging data, although often smaller than genomic data, can start to add up quickly with more images and samples.

Here is an table of average file sizes for various medical imaging modalities from @liu_imaging_2017:

```{r, fig.align='center', echo = FALSE, fig.alt= "Table of file types for imaging data, most modalities have files in the range of MB to GB. Note that these are approximate values.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_35")
```
[[source](https://www.mdpi.com/2078-2489/8/4/131)]


Note that depending on the study requirements, several images may be needed for each sample. Thus data storage needs can add up quickly.

```{r, fig.align='center', echo = FALSE, fig.alt= "Example table of overall file storage needs for samples in imaging studies.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_25")
```


### Clinical Data File Sizes

Really large clinical datasets can also produce sizable file sizes. For example the [Healthcare Cost and Utilization Project (HCUP) National (Nationwide) Inpatient Sample (NIS)](https://www.hcup-us.ahrq.gov/db/nation/nis/nisdbdocumentation.jsp) contains data on more than seven million hospital stays in the United States with regional information.

According to the NIS website it "enables analyses of rare conditions, uncommon treatments, and special populations" [@NIS].

Looking at the [file sizes](https://www.hcup-us.ahrq.gov/db/state/sedddist/sedddist_filesize.jsp) for the NIS data for different states across years, you can see that there are files for some states, such as California as large as 24,000 MB or 2.4 GB [@NIS]. You can see how this could add up across years and states quite quickly.


```{r, fig.align='center', echo = FALSE, fig.alt= "Table of file sizes for the Healthcare Cost and Utilization Project (HCUP) National (Nationwide) Inpatient Sample (NIS) of data from different years and states.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_42")
```

### Checking file sizes on Mac

If you own a Mac and want to check the size of a particular file, look at your file within a finder window. You can open a new finder window by clicking on the button that looks like a square with two colors and a face, typically in the bottom left corner if your dock or the strip of icons on your screen to help you navigate to different application programs.

```{r, fig.align='center', echo = FALSE, fig.alt= "Mac finder button", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_120")
```
Once you open a finder window, you can navigate to one of your files.

If you have the view setting that looks like 4 lines, you will get information about the size of each file.
```{r, fig.align='center', echo = FALSE, fig.alt= "Finder windows show file sizes on a Mac", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_133")
```



You can right click on  a file and click the "Get Info" button. This will give your more specific information.

```{r, fig.align='center', echo = FALSE, fig.alt= "Right clicking on a file in the finder window can give you more info about a file.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_128")
```


### Checking file sizes on PC/Windows

In a similar manner to checking file sizes on a Mac, with a Windows or PC computer, you can navigate to files by first opening the File Explorer application by typing this in the search bar next to the "start" button.

```{r, fig.align='center', echo = FALSE, fig.alt= "Finding the File Explorer on a Windows/PC computer", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_139")
```


Then navigate to a file of interest which will show information about the size in one of the columns to the right, if you hover over the file name, you will get more specific information.

```{r, fig.align='center', echo = FALSE, fig.alt= "File sizes are listed for each file within File Explorer windows on Windows/PC computers", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_145")
```


## Computing Options

### **Personal computers**

These are computers that your lab might own, such as a laptop, a desktop, used by one individual or maybe just a few individuals in your lab.  

```{r, fig.align='center', echo = FALSE, fig.alt= "A computer chip is also called the CPU. Inside this CPU or chip  are often multiple cores, which are actually individual CPUs.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_3")
```


If you are not performing intensive computational tasks, it is possible that you will only need personal computers for your lab. However, you may find that this changes, and you might require connecting your personal computers to shared computers for more computational power and or storage.


### **Shared Computing Resources**

What if you decide that you do need more computational power than your personal computer? You may encounter times where certain informatics tasks take way too long or are not even possible. Evaluating the potential file sizes of the data that you might be working with is a good place to start. However, keep in mind that sometimes certain computations may require more memory than you expect. This is particularly true when working with genomic or image files which are often compressed. So what can you do when you face this issue?

One great option, which can be quite affordable is using a server.

In terms of hardware, the term [server](https://techterms.com/definition/server) means a computer (often a computer that has much more storage and computing capacity than a typical computer) or groups of computers that can be accessed by other computers using a local network or the internet to perform computations or store data [@server_def]. They are often shared by people, and allow users to perform more intensive computational tasks or store large amounts of data. Read [here](https://en.wikipedia.org/wiki/Server_(computing)) to learn more [@server_2021].

Using a group of computers is often a much more cost effective option than having one expensive supercomputer (a computer that individually has the computational power of many personal computers) to act as a server [@supercomputer_2022]. It turns out that buying several less powerful computers is cheaper. In some cases however, an institute or company might even have a sever with multiple supercomputers!


As an example use of a server, your lab members could connect to a server from their own computers to allow each of them more computational power. Typically computers that act as servers are set up a bit differently than our personal computers, as they do not need the same functionality and are designed to optimize data storage and computational power. For instance they often don't have capabilities to support a [graphical user interface](https://www.omnisci.com/technical-glossary/graphical-user-interface) (meaning the visual display output that you see on your personal computer) [@GUI].



Instead they are typically only accessed by using a [command-line interface](https://en.wikipedia.org/wiki/Command-line_interface), meaning that users write code instead of using buttons like they might for a program like Microsoft Word that uses a graphical user interface [@command-line_2022]. In order to support this they have memory, processors or CPUs, and storage like your laptop.

Here is what a server might look like:

```{r, fig.align='center', echo = FALSE, fig.alt= "A physical server is a group of computers somewhere that connects to other computers. Such a server looks something like this.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_58")
```

In this case we have a group of computers making up this server. Here we see the nodes (the individual computers that make up the server) stacked in columns.


Among shared computing resources/servers there are three major options:

* **Clusters** - institutional or national resources
* **Grids** - institutional or national resources
* **Cloud** - commercial or national resources


### Computer Cluster

In a [computing cluster](https://en.wikipedia.org/wiki/Computer_cluster) several of the **same** type of computer (often in close proximity and connected by a local area network with actual cables or an [intranet](https://www.igloosoftware.com/blog/internet-vs-intranet-vs-extranet-whats-the-difference/) rather than the internet) work together to perform pieces of the same single task simultaneously [@computer_cluster_2022]. The idea of performing multiple computations simultaneously is called [parallel computing](https://en.wikipedia.org/wiki/Parallel_computing) [@parallel_2021].

There are different designs or architectures for clusters. One common one is the [Beowulf cluster](https://en.wikipedia.org/wiki/Beowulf_cluster) in which a master computer (called front node or server node) breaks a task up into small pieces that the other computers (called **client nodes** or simply **nodes**) perform [@beowulf_2022].

For example, if a large file needs to be converted to a different format, **pieces** of the file will be converted simultaneously by the different nodes. Thus each node is performing the **same task** just with different pieces of the file. The user has to write code in a special way to specify that they want parallel processing to be used and how. See [here](https://www.freecodecamp.org/news/how-to-supercharge-your-bash-workflows-with-gnu-parallel-53aab0aea141/) for an introduction about how this is done @Zach_Caceres_GNU_Parallel_2019.

It is important to realize that the CPUs in each of the node computers connected within a cluster are all performing a similar task simultaneously.

See [here](https://cs.wmich.edu/~elise/courses/cs626/s09/hussein/Parallel_and_Cluster_Computing.pdf) for more information [@de_doncker].

### Computer Grid

In a [computing grid](https://hazelcast.com/glossary/grid-computing/) are often **different** types of computers in **different** locations work towards an overall common goal by performing **different** tasks [@grid].

Again, just like computer clusters, there are many types of architectures that can be rather simple to very complex. For example you can think of different universities collaborating to perform different computations for the same project. One university might perform computations using gene expression data about a particular population, while another performs computations using data from another population. Importantly each of these universities might use clusters to perform their specific task.

Both grids and clusters use a special type of software called middleware to coordinate the various computers involved.
Users need to write their scripts in a way that can be performed by multiple computers simultaneously. Users also need to be conscious of how to schedule their tasks and to follow the rules and etiquette of the specific cluster or grid that they are sharing (more on that soon!).

See [here](https://pediaa.com/difference-between-cluster-and-grid-computing/)  and [here](https://www.geeksforgeeks.org/difference-between-grid-computing-and-cluster-computing/)for more information about the difference between clusters and grids [@lithmee_difference_2018;  @grid_cluster_difference_2019].


### "Cloud" computing

More recently, the ["Cloud"](https://en.wikipedia.org/wiki/Cloud_computing) has become a common computing option. The term "cloud" has become a widely used buzzword [@cha_cloud_2015] that actually has a few slightly different definitions that have changed overtime, making it a bit tricky to keep track of. However, the "cloud"  is typically meant to describe large computing resources that involve the connection of **multiple servers** in multiple locations to one another [@cloud_2022] using the internet. See [here](https://www.redhat.com/en/topics/cloud-computing/cloud-vs-virtualization) for a deeper description of what the term cloud means today and how it compares to other more traditional shared computing options [@cloud_deeper].

Many of us use cloud storage regularly for Google Docs and backing up photos using iPhoto and Google. Cloud computing for research works in a similar way to these systems, in that you can perform computations or store data using an available server that is part of a larger network of servers. This allows for even more computational dependability beyond a more simple cluster or grid. Even if one or multiple servers is down, you can often still use the other servers for the computations that you might need.

Furthermore, this also allows for more opportunity to scale your work to a larger extent, as there is generally more computing capacity possible with most cloud resources [@cloudvstrad].


Companies like Amazon, Google, Microsoft Azure, and others provide cloud computing resources. **Somewhere these companies have clusters of computers that paying customers use through the internet.**  In addition to these commercial options, there are newer national government funded resource options like [Jetstream](https://portal.xsede.org/jetstream) (described in the next section).  We will compare computing options in another chapter coming up.



```{r, fig.align='center', echo = FALSE, fig.alt= "Cloud computing options", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_28")
```


### Accessing Shared Computer Resources

It's important to remember that all of the shared computing options that we previously described involve a [data center](https://en.wikipedia.org/wiki/Data_center) where are large number of computers are physically housed.

```{r, fig.align='center', echo = FALSE, fig.alt= "Examples of servers or shared computers include clusters  that may exist at your institution or national computing resources like Xsede.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_23")
```


You may have access to a [HPC (which stands for High Performance Computing) cluster](https://www.hpc.iastate.edu/guides/introduction-to-hpc-clusters/what-is-an-hpc-cluster) at your institute. This can be a great cost-effective and typically secure option.  


If your university or institution has a HPC [cluster](https://en.wikipedia.org/wiki/Computer_cluster), this means that they have a group of computers acting like a server that people can use to store data or assist with intensive computations. Often institutions can support the cost of many computers within an HPC cluster. This means that multiple computers will simultaneously perform different parts of the computing required for a given task, thus significantly speeding up the process compared to you trying to perform the task on just your computer!


If your institute doesn't have a shared computing resource like the HPCs we just described, you could also consider a national resource option like [Xsede](https://www.xsede.org/).
[Xsede](https://www.xsede.org/) is led by the University of Illinois National Center for Supercomputing Applications (NCSA) and includes 18 other partnering institutions (which are mostly other universities). Through this partnership, they currently support 16 supercomputers. Universities and non-profit researchers in the United States can request access to their computational and data storage resources. See [here](https://portal.xsede.org/allocations/resource-info) for descriptions of the available resources.


Here you can see a photo of Stampede2, one of the supercomputers that members of Xsede can utilize.

```{r, fig.align='center', echo = FALSE, fig.alt= "An image of Stampede2 one of the supercomputers that members of Xsede can use.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf9c252d058_0_63")
```


[[source](https://www.xsede.org/ecosystem/resources)]

> Stampede2, generously funded by the National Science Foundation (NSF) through award ACI-1134872, is one of the Texas Advanced Computing Center (TACC), University of Texas at Austin's flagship supercomputers.

See [here](https://portal.xsede.org/tacc-stampede2) for more information about how you could possibly connect to and utilize Stampede2.

Importantly when you use shared computers like national resources like Stampede2 available through Xsede, as well as institutional HPCs, you will share these resources with many other people and so you need to learn the proper etiquette for using and sharing these resources. We will discuss this more in a coming chapter.

However, there is also now an option to access the different XSEDE computing resources through a cloud environment option called [Jetstream2](https://jetstream-cloud.org/).  

Here is a video about Jetstream2:

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/NQ3flxJANTw")
```




We will also discuss how the use of these various computing options differ in the next chapters. Importantly there are also some computing platforms that have been especially designed for scientists and specific types of researchers, so it is also useful to know about these options.


## Conclusion

We hope that this chapter has given you some more perspective on how large medical research data files can be, as well as given you more familiarity with how well your computer might be able to accommodate the files that you might work with. We also hope that this chapter has provided you with some more awareness about computing options that might be available to you, should you need more capacity than your current computer.


# Shared Computing Etiquette

In this chapter we will discuss the proper etiquette for using more traditional shared computing resources, such as an institutional high performance computing cluster server. This will help you to understand what would be required for you to use such resources. Different resources will have slightly different use rules, however, many resources will share common usage requirements. The following is written based on personal experience, the @doi_rules and the @JHPCE.


```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1. Describe why traditional shared computing resources need usage rules, 2. Define common best practices for proper etiquette on traditional shared computing resources", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.g11383d0152c_0_12")
```

We will use the Johns Hopkins Joint High Performance Computing Exchange (JHPCE) cluster resource as an example to motivate the need for usage rules and proper sharing etiquette for such resources.

First let's learn a bit about this JHPCE. For this particular resource there are about 400 active users.It is optimized for genomic and biomedical research and has 4,000 cores! That's right, as you can imagine, this is much more powerful than the individual laptops and desktops that researchers at the university have for personal use, which would typically currently only have around 8 cores. There is also 28TB of RAM and 14 PB of storage!

Now that you know more about digital sizes, you can appreciate that this server can allow for much faster processing and really large amounts of storage, as again a researchers' computer might have something like 16 GB of RAM and 1TB of storage.

There are 68 nodes that make up the JHPCE currently. As, with most clusters some of the nodes are dedicated to managing users logging in to the cluster and some of the nodes are dedicated to data transferring. Each node has 2-4 CPUs that provide 24-128 cores! As you can see these processors or chips have a lot more cores per each CPU than a typical personal computer.

Individual users connect and perform jobs (aka computational tasks) on the cluster using a formal [common pool resource (CPR)](https://en.wikipedia.org/wiki/Common-pool_resource) hierarchy system. What does this mean? This means that it is a shared resource, where if one user overused the resource it would be to the detriment of others and to overcome this there are usage rules and regulations that are enforced by managers of the resource @common-pool_2022.  This is important because if a single or a few users used up all the computing resources one day, then the other nearly 400 users would have to delay their work that day, which would not be fair.

## General Guidelines for shared computing resources

Each cluster or other shared computing resource will have different rules and requirements, but here are a few general rules to keep in make sure that you don't accidentally abuse the privilege of sharing an amazing resource like this. Don't be too worried, most shared resources will give you guidance about their specific rules and will often also have settings that don't allow users to make major blunders.

### Security guidelines

One major aspect to consider is keeping the computers in the cluster safe from harm. You wouldn't want to lose your precious data stored on the cluster and neither would your colleagues!

 - Use a good [secure password](https://its.lafayette.edu/policies/strongpasswords/) that is not easy for someone else to guess.

 Some people suggest using sentences that are easy for you to remember, you could consider a line of lyrics from song or poem that you like, or maybe a movie. Modify part of it to include symbols and numbers [@passwords].


```{r, fig.align='center', echo = FALSE, fig.alt= "Cartoon - One character says:I came up with the best password! It’s from Elton John’s Rocket Man song. And all this science I don’t understand It’s just my job five days a week A rocket man A rocket man. All the Os are zeros. The other character says: That is super good! But remember you aren’t supposed to share your passwords. Plus now all the readers of this chapter know too.The original character says: Oh shoot… you’re right. The other character says: It’s OK, sounds like you can come up with another.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.g11383d0152c_0_22")
```


 - Don't share your password and keep it safe!

 If you have a Mac, you could consider storing it in your [Keychain](https://support.apple.com/en-ie/guide/mac-help/mchlf375f392/mac), alternatively if you have a different type of computer or don't like the Mac Keychain, consider [Dashlane](https://www.dashlane.com/) or other password manger services. Luckily both of these options do not come at any extra cost and can be helpful for storing all the passwords we use regularly safely. These are especially good options if your password is difficult for you to remember. Make sure that you abide by any rules regarding storing passwords that might be required by the resource you intend to use.


```{r, fig.align='center', echo = FALSE, fig.alt= "Cartoon - One character says: Hey, what do you have there?. The other character says: Oh just bringing my passwords with me in case I forget. I’ve secured them carefully on paper with invisible ink, in a cypher with its own code, inside a fireproof box with a lock. The original character says: That’s very impressive. You could also just use a password manager. The other character says: Oh that might be good… because this fireproof box is quite heavy!", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.g10e2895be5b_58_54")
```


 - Don't access a server on a computer that is not authorized to do so.

Some servers will require that your computer be authorized for access for added security. It's a good idea to follow these rules. If you can, perhaps authorize a laptop in case you might need to gain access when you need to be out of town. However if you do so, make sure you also only access such servers with a secure WiFi network. One way to ensure this is is to avoid using public WiFi networks. If you must use a public WiFi network, consider using a [virtual private network (VPN)](https://en.wikipedia.org/wiki/Virtual_private_network) for added security. Here is an [article](https://www.wired.com/story/best-vpn/) about different VPN options [@gilbertson_4_2021].

 -  Do not alter security settings without authorization.

Loosening security settings could pose a risk to the data stored on the server. On the other hand, making more strict security settings could cause other users to not be able to perform their work. Contact the managers of the resource if you think changes need to be made.

 - Immediately report any data security concerns.

To protect the integrity of your data and your colleagues, be sure to report anything strange about the shared computing resource to those who manage it so that they can address it right away. Also report to them if you have any security breaches on the computer(s) that you use to access the shared computing resource.

### Overall use guidelines

Now that we know how to keep the resource safe, let's next talk about general usage.

 - Don't install software unless you have permission.

It is possible that the software you want to use might already be installed somewhere on the shared computing resource that you are unaware about. In addition, if you install a different version of a software program, it is possible that this version (especially if it is newer) will get automatically called by other people's scripts. This could actually break their scripts or modify their results. They may have a reason to use an older version of that software, do not assume that they necessarily want the updated version. Instead, let the managers of the resource know. They can inform other users and make sure that everyone's work will not be disrupted.

 - Don't use the server for storage or computation that you are not authorized for.

This is often a rule for shared computing resources, simply because such shared resources are intended for a specific reason and likely funded for that reason. Such resources are costly, and therefore the computational power should be used only for what it is intended for, otherwise people may view the use of the resources for other purposes as essentially theft.

 - Don't alter configurations without authorization.

This could result unintended and unexpected consequences for other users.


### Daily use guidelines

Now let's discuss how you should use such resources on a daily basis.

When you submit jobs, make sure you follow the following guidelines. Again consider the fact that there may be more or different requirements for the specific resource that you might be using.

 - Don't use the login or transfer nodes for your computations.

 This will cause issues for other users in terms of logging in and transferring their data. This could cause them to be unable to do their work.

 - Think about memory allocation and efficiency.

Consider how much RAM and storage is available for people on the shared computing resource. Try not to overload the resource with a really intensive job or jobs that will use most of the resources and either slow down the efficiency of the work for others or not allow them to perform their work at all.

This involves:

   * Not using too many nodes if you don't need to
   * Not using too much RAM on a given node or overall if you don't need to
   * Not submitting too many jobs at once
   * Communicating with others to give them advanced warning if you are going to submit large or intensive jobs

If you have a really large job that you need to perform, talk with the managers of the resource so that you can work out a time when perhaps fewer users would be inconvenienced. Consult the guidelines for your particular resource about how one let's people know about large jobs before you email the administrators of the resource directly. Often their are communications systems in place for users to let each other know about large jobs.

### Communication Guidelines

Speaking of communication, let's dive into that deeper for a bit.

- Use the proper order for communication.

Often shared resources have rules about how they want people to communicate. For example for some resources it is suggested that you first ask your friends and colleagues if you are confused about something, then consult any available forums, if that does not work then directly email the administrators/managers of the resource. Keep in mind that these people are very busy and get lots of communications.

- Use the ticket system

If a resource has a ticket system for users to get support, use it instead of communicating by email. If such a system is in place, then the administrators running it are used to getting requests this way. If you email directly, you may not receive feedback in a timely manner or the email might get lost.

### Specific Rules

Ultimately it is very important to learn about the rules, practices, and etiquette for the resource that you are using and to follow them. Otherwise, you could lose access. Often other users are a great resource!


## Interacting with shared resources

Often you will need to use the command line to interact with a server from your personal computer. To do so on a Mac or a Linux computer you can typically do so using the terminal program that is already on your computer. For PC or Windows computer users, you can use programs like [MobaXterm](http://mobaxterm.mobatek.net/).

If you wish to run a program with a graphical interface, then you might need to have a program to help you do so. On Macs, you can download [XQuartz](http://xquartz.macosforge.org/landing/). If you use MobaXterm on your PC or Windows computer, then you will already be set. Linux computers also typically should already have what you need.

If you are new to Unix commands check out this cheat sheet below.

```{r, echo = FALSE, fig.alt= "Cheat sheet of Unix commands at https://files.fosswire.com/2007/08/fwunixref.pdf"}
knitr::include_url(url = "https://files.fosswire.com/2007/08/fwunixref.pdf")
```


## Running Jobs

Typically a program is used to schedule jobs. Remember that jobs are the individual computational tasks that you ask the server to run. For example, this could be something as simple as moving large files from one directory to another or as complex as running a complicated script on a file.

Such job scheduling programs assign jobs to available node resources as they become available and if they have the required resources to meet the job. These programs have their own commands for running jobs, checking resources, and checking jobs. Remember to use the management system to run your jobs using the compute nodes not the login nodes (nodes for users to log in). There are often nodes set up for transferring files as well.

In the case of the JHPCE, a program called Sun Grid Engine (SGE) is used, but there are others job management programs. See [here](https://jhpce.jhu.edu/wp-content/uploads/2021/06/JHPCE-Overview-2021-10.pdf) for more information on how people use SGE for the JHPCE shared resource.

### Specifying memory (RAM) needs

Often there is a default file size limit for jobs. For example the JHPCE has a 10GB file size limit for jobs. You may need to specify when you have a job using a file that exceeds the file size limit and set the file size for that job. As you may recall if you are using whole genome files you are likely to exceed the default file limit size. Often you are also given a default amount of RAM for your job as well. Again, you can typically run a job with more RAM if you specify. Similar to the file size limit, you will likely need to set the RAM that you will need for your job if it is above the default limit. Often this involves setting a lower and upper limit to the RAM that your job can use. If your job exceeds that amount of RAM it will be stopped. Typically people call stopping a job "killing" it. The lower and upper limit can be the same number.

How do you know how much RAM to assign to your job? Well if you are performing a job with files that are two times the size of the file size default limit, then it might make sense to double the RAM you would typically use. It's also a good idea to test on one file first if you are going to perform the same job on multiple files. You can then assess how much RAM the job used. First try to perform the job with lower limits and progressively increase until you see that the job was successful and not killed for exceeding the limit.  Keep in mind however how much RAM there is on each node. Remember, it is important to not ask for all the RAM on a single node or core on that node, as this will result in you hogging that node and other users will not be able to use RAM on that node or core on that node. Remember that you will likely have the option to use multiple cores, this can also help you to use less RAM across each core. For example, a job that needs 120GB of RAM could use 10 cores with 12 GB of RAM each.

Often there will be a limit for the number of jobs, the amount of RAM, and the number of cores that a single user can use beyond the default limits. This is to ensure that a user doesn't use too many resources causing others to not be able to perform their jobs. Check to see what these limits are and then figure out what the appropriate way is to contact to request for more. Again communication standards and workflows may vary based on the resource.

### Checking status

 It's also a good idea to check the status of your jobs to see if they worked or got killed. You can check for the expected file outputs or there are commands for the server management software that can help you check currently running jobs.

## Storage

Often you will be given a home directory which will likely be backed up, however, other storage directories often will not be. Be careful about where you store your data, as some directories might be for temporary use and get wiped to keep space available for others.

## Research Platforms

In this chapter we will provide examples of computing platforms that are designed to help researchers and that you might find useful for your work. Please note that we aim to provide a general overview of options and thus this is not a complete list. Let us know if there is a platform or system that you think we should include!

<div class = "warning">

 We highly suggest you also **read the next chapter**, which will point out important considerations to think about when deciding to work on a shared computing resource platform like those discussed in this chapter.

</div>


```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1.Recognize general informatics computing and data management platforms. 2.Be aware of differnt benefits of the various platforms", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf5f8818810_1_171")
```

The major advantage of these platforms is that users can analyze data where it lives, as many platforms host public data. However, some also allow you to upload your own data.  There is less need for data transfers back and forth to your personal computer, as you can analyze your data, store your data and share it in one place, saving time. Users can sometimes also share how they did their analysis as well, improving reproducibility practices. Additionally, another advantage is that some of these platforms also provide educational material on how to work with data.

Many offer a [graphical user interface](https://www.omnisci.com/technical-glossary/graphical-user-interface) also simply called just graphical interface or GUI, allows for users to choose functions to perform by interacting with visual representations, which can be useful for individuals how are less comfortable writing code. They have a "user-centered" design that creates a visual environment where users can for example **click on** tabs, boxes, or icons for to perform functions. This also often allows users to more directly see plots and other types of visualizations.

Some platforms also offer a [command line interface](https://searchwindowsserver.techtarget.com/definition/command-line-interface-CLI) (also known as a character interface) which allows for software functions to be performed by specifying through commands written in text. This typically offers more control than a GUI, however command line interfaces are often less user friendly as they require that the user know the correct commands to use.


### National Cancer Institute Cloud Resources

Funded by the [National Cancer Institute (NCI)](https://www.cancer.gov/), the [cancer research data commons](https://datascience.cancer.gov/data-commons) provides data access and computing infrastructure for researchers through [three different platforms](https://dcf.gen3.org/cloud-resources), the [Cancer Genomics Cloud (CGC)](https://www.cancergenomicscloud.org/) which uses Google Cloud resources, the [Institute for Systems Biology Cancer Gateway in the Cloud (ISB-CGC)](https://isb-cgc.appspot.com/) which also uses Google Cloud resources, and [FireCloud](https://portal.firecloud.org/) from the Broad Institute, which uses Amazon Cloud resources.

### Cancer Genomics Cloud

The [Cancer Genomics Cloud (CGC)](https://www.cancergenomicscloud.org/) is a computing platform that researchers can used to analyze, store, and share their own data, as well as work with large public and controlled cancer data sets, including genomic and imaging data. CGC offers tutorials and guides to help research get started, as well as $300 of free credits to use the platform and test it out. Users can also access many tools and workflows to help them perform there analyses. CGC also offers regular [webinars](https://www.cancergenomicscloud.org/webinars).

The platform is based on a partnership with [Seven Bridges](https://www.sevenbridges.com/), a biomedical analytics company, and can be accessed simply by using a web browser. Users can can use a point and click system also called a graphical user interface (GUI) or can access resources using the command line. See this [link](https://www.cancergenomicscloud.org/getting-started) to learn more.


### Institute for Systems Biology (ISB) Cancer Gateway in the Cloud

The [ISB-CRC](https://isb-cgc.appspot.com/) platform allows users to browse and data from the [Genomic Data Commons](https://gdc.cancer.gov/) and other sources, including sequencing and imaging data both public and controlled. They provide access pipeline tools, as well as to pipelines, workflows, and Notebooks written by others in R and Python to help users perform analyses. ISB also offers $300 in [free credits](https://isb-cancer-genomics-cloud.readthedocs.io/en/latest/sections/HowtoRequestCloudCredits.html) to try out the platform. See [here](https://isb-cgc.appspot.com/how_to_discover/#0) for a user guide.


### Broad Institute FireCloud

[FireCloud](https://portal.firecloud.org/) provides users with computing resources and access to workspaces using Broad's tools and pipelines. Users can run large scale analyses and work with collaborators. FireCloud offers access to [The Cancer Genome Atlas (TCGA)](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga) controlled-access data. Other platforms like Galaxy and Terra described next, share resources with FireCloud.

### Galaxy

This section was written by [Jeremy Goecks](https://goeckslab.org/people/jeremy.html):

Galaxy is a web-based computational workbench that connects analysis tools, biomedical datasets, computing resources, a graphical user interface, and a programmatic API. Galaxy (https://galaxyproject.org/) enables accessible, reproducible, and collaborative biomedical data science by anyone regardless of their informatics expertise. There are more than 8,000 analysis tools and 200 visualizations integrated into Galaxy that can be used to process a wide variety of biomedical datasets. This includes tools for analyzing genomic, transcriptomic (RNA-seq), proteomic, metabolomic, microbiome, and imaging datasets, tool suites for single-cell omics and machine learning, and thousands of more tools. Galaxy’s graphical user interface can be used with only a web browser, and there is a programmatic API for performing scripted and automated analyses with Galaxy.

```{r, fig.align='center', echo = FALSE, fig.alt= "Galaxy can be accessed through a web browser and provides users with access to tools, datasets, computing resources, a graphical user interface (GUI) for users who would like to interact with Galaxy by clicking buttons and using drop-down menus and a programmtic API for users that would like to write code to interact with Galaxy", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_131")
```

Galaxy is used daily by thousands of scientists across the world. A vibrant Galaxy community has deployed hundreds of Galaxy servers across the world, including more than 150 public and three large national/international servers in the United States, Europe, and Australia (https://usegalaxy.org, https://usegalaxy.eu, https://usegalaxy.org.au). The three national/international servers have more than 250,000 registered users who execute >500,000 analysis jobs each month. Galaxy has been cited more than 10,000 times with >20% from papers related to cancer. The Galaxy Tool Shed (https://usegalaxy.org/toolshed) provides a central location where developers can upload tools and visualizations and users can search and install tools and visualizations into any Galaxy server. Galaxy has a large presence in the cancer research community. Galaxy serves as an integration and/or analysis platform for 7 projects in the NCI ITCR program. There is also increasing use of Galaxy in key NIH initiatives such as the NCI Cancer Moonshot Human Tumor Atlas Network (HTAN) and the NHGRI Data Commons, called the AnVIL (https://anvilproject.org/).

```{r, fig.align='center', echo = FALSE, fig.alt= "Map of the 3 Galaxy servers", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_135")
```


Galaxy’s user interface, accessible via a web browser, provides access to all Galaxy functionality. The main Galaxy interface has three panels: available tools (left), running analyses and viewing data (middle), and a full history of tools run and datasets generated (right).

```{r, fig.align='center', echo = FALSE, fig.alt= "Galaxy user interface showing the 3 panels: left = tools, middle = analysis and data, and right = history and generated datasets", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_111")
```


Datasets for analysis in Galaxy can be **uploaded** from a laptop or desktop computer or obtained from public data repositories connected to Galaxy. With Galaxy, complex workflows composed of tens or even hundreds of analysis tools can be created and run. In Galaxy’s workflow interface, tools can be added and connected via a simple drag-and-drop approach.

```{r, fig.align='center', echo = FALSE, fig.alt= "Galaxy workflow example showing the beginning steps of a workflow", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_141")
```


Galaxy users can share all their work—analysis histories, workflows, and visualizations—via simple URLs that are available to specific colleagues or a link that anyone can access. Galaxy’s user interface is highly scalable. Tens, hundreds, or even thousands of datasets can be grouped into collections and run in parallel using individual tools or multi-tool workflows. In summary, Galaxy is a popular computational workbench with tools and features for a wide variety of data analyses, and it has broad usage in cancer data analysis.

See [here](https://toolshed.g2.bx.psu.edu/) for the list of applications supported by Galaxy and [here](https://training.galaxyproject.org/) for more information on how to use Galaxy resources.


### Terra

[Terra](https://terra.bio/) is a biomedical research computing platform that is based on the Google Cloud platform, that also allows users easier ways to manage the billing of their projects. It provides users with access to data, workflows, interactive analyses using Jupyter Notebooks, RStudio, and Galaxy, data access and tools from [FireCloud from the Broad Institute](https://firecloud.terra.bio/), as well as workspaces to organize projects and collaborate with others. Terra also has [many measures](https://terra.bio/resources/security/) to help ensure that data is secure and they offer clinical features for ensuring that [health data is protected](https://terra.bio/about/privacy/). Note that users who do upload protected health information must select to use  extra clinical features and enter a formal agree with [Terra/FireCloud](https://firecloud.terra.bio/) about their data. See [here](https://support.terra.bio/hc/en-us/articles/360024688731-Terms-of-Service) for more information.

Importantly users can get access to use [Genotype -Tissue Expression (GTEx)](https://gtexportal.org/home/), [Therapeutically Applicable Research to Generate Effective Treatments (TARGET)](https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000218.v24.p8) and [The Cancer Genome Atlas (TCGA)](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga) data using the platform. See [here](https://support.terra.bio/hc/en-us/articles/4402326091675-Accessing-GTEx-TARGET-TCGA-data) for information on how.

Users can pay for data storage and computing costs for Google Cloud through Terra. Users can browse data for free.

Check out this video for more information:

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/3rH86vcAqK8")
```



## AnVIL

If you could use some guidance on how to perform analyses using Galaxy and Terra, especially for genomic research, check out [AnVIL](https://anvilproject.org/), the [National Human Genome Research Institute (NHGRI)](https://www.genome.gov/) Analysis Visualization and Informatics Lab-space. It also provides access to many important genomic and related [datasets](https://anvilproject.org/data) from the NHGRI.


According to their website:

> By providing a unified environment for data management and compute, AnVIL eliminates the need for data movement, allows for active threat detection and monitoring, and provides elastic, shared computing resources that can be acquired by researchers as needed.

It relies on Terra for the cloud based compute environment, Dockstore for  standardized tools and workflows, Gen3 for data management for querying and organizing data, Galaxy tools and environment for analyses with less code requirements, and [Bioconductor](https://www.bioconductor.org/) tools for R programming users. [Bioconductor](https://www.bioconductor.org/) is a project with the mission to catalog, support, and disseminate bioinformatics open-source R packages. Packages have to go through a review process before being included.


## CyVerse

[CyVerse](https://cyverse.rocks/about) is a  similar computing platform that also offers computing resources for storing, sharing, and working with data with a graphical interface, as well as an API. Computing was previously offered using the cloud computing platform from CyVerse called [Atmosphere](https://cyverse.org/refocusing-atmosphere-to-support-cloud-native-development), which relied on users using virtual machines. Users will now use a new version of Atmosphere with partnership with [Jetstream](https://jetstream-cloud.org/). This allows users to use containers for easier collaboration and also offers US users more computing power and storage. Originally called iPlant Collaborative, it was started by a funding from the National Science Foundation (NSF) to support life sciences research, particularly to support ecology, biodiversity, sustainability, and agriculture research. It is led by the University of Arizona, the Texas Advanced Computing Center, and Cold Spring Harbor Laboratory. It offers access to an environment for performing analyses with Jupyter (for Python mostly) and RStudio (for R mostly) and a variety of tools for Genomic data analysis. See [here](https://cyverse.atlassian.net/wiki/spaces/DEapps/pages/241882146/List+of+Applications) for a list of applications that are supported by CyVerse.  Note that you can also install tools on both platforms. Both CyVerse and Galaxy offer lots of helpful documentation, to help users get started with informatics analyses.

See [here](https://learning.cyverse.org/en/latest/) to learn more.

```{r, fig.align='center', echo = FALSE, fig.alt= "CyVerse graphical interface for performing analyses", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd56752f25_0_0
")
```

## SciServer

SciServer is accessible through a web browser and allows users to store, upload, download, share, and work with data and common tools on the same platform. It was originally built for the astrophysics community (and called SkyServer) but it has now been adapted to be used by scientists of all fields and is indeed used by many in the genomics field.  It allows users to use Python and R in environments like Jupyter notebooks and RStudio, and also supports (Structured Query Language) SQL for data querying and management and is built on the use of Docker.

The main idea of SciServer, is based on this premise: "bring the analysis to the data". It is free to use after users register. However, users can buy extra resources. Users can keep data private or share their data.

As compared to Galaxy, this resources may be better for users with a bit more familiarity with informatics but who require more flexibility, particularly for working with collaborators such as physicists or material scientists as there are more tools supported across disciplines. In addition it also gives users access to very large data sets on Petabyte-scale (note that some of these require special permission to use) and supports developers to create their own web interfaces called SciUIs for particular use cases.

For @sciserver_2020 for more information.  


```{r, fig.align='center', echo = FALSE, fig.alt= "The SciServer dashboard allows users to interact with the various resources and tools available through SciServer. Most are available through the command line but some have user interface options. Jupyter Notebooks, RStudio and SciUIs (interfaces created by other SciServer users) can be used to interactively explore and analyze data by users.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd56752f25_0_13")
```



## Materials Cloud

Another resource that might be of interest to Python users, particular those who collaborate with material scientists, is Materials Cloud. It is designed to promote reproducible work, collaboration, and sharing of resources among scientists, particularly for simulations for the materials science field. Users can share data in a citable way, download data, upload data, share workflows, and perform analyzes.

```{r, fig.align='center', echo = FALSE, fig.alt= "Materials Cloud resources are based on allowing users to Learn about resources, Work using the resources, Discover aspects about data that is available, Explore data with interactive graphs, and archive to store and share data.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd56752f25_0_6")
```

This resource uses [AiiDAlab](https://www.materialscloud.org/work/aiidalab) as the computing environment for researchers, which is based on [AiiDA](https://www.nature.com/articles/s41597-020-00638-4). According to their website:

> AiiDAlab builds on [AiiDA](https://www.nature.com/articles/s41597-020-00638-4) as the computational workflow engine, and the Jupyter environment (notebooks, widgets, …) for writing and sharing apps

See [here](https://www.sciencedirect.com/science/article/pii/S092702562030656X?via%3Dihub) to learn more about how AiiDAlab supports the sharing of scientific workflows, particularly for those that use Python.

To learn more about Materials Cloud, check out @talirz_materials_2020.


## Overture

Overture is a relatively new option for perform large-scale genomic data analyses. You can upload, download, manage, analyze and share your data with authentication and authorization methods to add security. Although designed for genomic research, the [data management system](https://www.overture.bio/documentation/dms/) can be used for other scientific domains. Currently, additional products are still being developed for analysis, visualization, and sharing. However, several collaborations have created new incredible resources using some of the existing and developing products that might be useful for your research. Alternatively, Overture has options to help you create your own platform, see [here](https://www.overture.bio/services/) for more information. It is compatible with Google, Microsoft Azure, and PostgreSQL for storage options.

These collaborations using Overture products can be found on the [case studies](https://www.overture.bio/case-studies/) page of the [Overture website](https://www.overture.bio/).

For example, the [Cancer Genome Collaboratory](https://cancercollaboratory.org/) is one such collaboration. This is A cloud-based resource that allows researchers to perform analyses using [International Cancer Genome Consortium (ICGC)](https://en.wikipedia.org/wiki/International_Cancer_Genome_Consortium) cancer genome data, which includes tumor mutation data from the [The Cancer Genome Atlas (TCGA)](https://en.wikipedia.org/wiki/The_Cancer_Genome_Atlas) and the [Pan-Cancer Analysis of Whole Genomes (PCAWG)](https://dcc.icgc.org/pcawg) mutation data. See [here](https://cancercollaboratory.org/services-cloud-resources) for information about billing, storage capacity, access, and security.

In addition, Overture products have also been used to create other data resources, such as the [Kids First Data Resource Portal](https://portal.kidsfirstdrc.org/login)  which has childhood cancer and birth defect genomic data for over 76,000 samples, and the [National Cancer Institute's Genomic Data Commons Data portal](https://portal.gdc.cancer.gov/), which also includes [The Cancer Genome Atlas (TCGA)](https://en.wikipedia.org/wiki/The_Cancer_Genome_Atlas) and [Therapeutically Applicable Research to Generate Effective Treatments (TARGET)](https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000218.v24.p8). The portal supports some basic [analyses](https://portal.gdc.cancer.gov/analysis) as well for clinical data statistics and survival analysis.

## Globus

Globus provides developers of new platforms to manage, transfer, and share data with special attention to privacy and security.

It has been used for several platforms such as the  [Systems Biology Knowledgebase (KBase)](https://www.kbase.us/), which is focused on integrating data across plants and microbes, [Biomedical Research Informatics Network (BIRN)](https://en.wikipedia.org/wiki/Biomedical_Informatics_Research_Network), which is a collaborative project to bring biomedical researchers together and share resources and data [@helmer_enabling_2011], and [Globus Genomics](https://www.globusgenomics.org/genomics/about.html),which uses the Globus data management infrastructure, Amazon web services, and Galaxy workflows to assist researchers with their genomics research endeavors. See this [link](https://www.globus.org/user-stories) for more examples of how others have used Globus.


## BaseSpace Sequence Hub

[BaseSpace](https://basespace.illumina.com/) is a platform that allows for data analysis of Illumina sequencing data and syncs easily with any Illumina sequencing machines that you might work with. There are many [applications](https://www.illumina.com/products/by-type/informatics-products/basespace-sequence-hub/apps.html) available to help you with your genomics research. They offer a 30 day free trial.


## ATLAS.ti

[ATLAS.ti](https://atlasti.com/) is designed particularly for qualitative analysis. You can use a variety of data types including video, audio, images, surveys, and social media data. A variety of tools, particularly for text data analysis are provided for methods such as [sentiment analysis](https://en.wikipedia.org/wiki/Sentiment_analysis), which is the process of assigning a general tone or feeling to text and [named-entity recognition](https://en.wikipedia.org/wiki/Named-entity_recognition), which is the process of extracting certain characteristics from texts that are what is called a [named entity] or a real-world object - such as a person's name or address. Such analyses can be helpful for understanding behaviors that might be associated with cancer risk. Although this type of analysis can be performed using R or Python among other coding languages, ATLAS.ti offers a nice graphical user interface to perform these types of analyses.Furthermore ATLAS.ti offers a great deal of flexibility about such analyses using different data types easily.

```{r, echo = FALSE, fig.alt= "Cheat sheet of Unix commands at https://files.fosswire.com/2007/08/fwunixref.pdf"}
knitr::include_url(url = "https://downloads.atlasti.com/docs/branding/atlasti_brochure_v9_EN_interactive_202110.pdf")
```



## GenePattern

[GenePattern](https://www.genepattern.org/) is similar to Galaxy in that it provides a web-based interface for genomic analyses. You can upload your own data, use workflows and pipelines form others and more!

See [here](https://www.genepattern.org/user-guide) to access their user guide and [here](https://notebook.genepattern.org/quickstart/) for a quick start guide to using [GenePattern Notebook](https://notebook.genepattern.org/) which uses [Jupyter Notebooks](https://jupyter.org/) and GenePattern analysis tools to easily create data analysis reports. Users can also publish and share their notebooks with collaborators or the field, as well as access other people's notebooks that they can adapt for their own uses. See [here](https://notebook.genepattern.org/library/) for a collection of available notebooks.

## XNAT

[XNAT](https://www.xnat.org/about/) offers computing resources and tools for performing imaging analysis and for storing and sharing imaging data in a HIPAA complaint manner (more on that in the coming). Developed by the [Bukner lab](https://cnl.rc.fas.harvard.edu/) previously at the Washington University and now at Harvard, it supports a variety of imaging data as well as other data types like clinical data.  Some tools can be used with a graphical interface and others with the command-line. See [here](https://wiki.xnat.org/documentation/case-studies) for example use cases. There is also a great deal of documentation available about how to use the tools and resources available at https://wiki.xnat.org/documentation.

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/ENk589mOkhI")
```


## OHIF

The [open health imaging foundation (OHIF)](https://ohif.org) is a web-based imaging analysis platform that is widely used, particularly for radiology analysis, but it also supports whole-slide microscopy image analysis. It was developed by [Gordon Harris](https://ohif.org/team/) et al. and can be used for a variety of applications from cardiology to veterinary medicine. Check out these [example use cases](https://ohif.org/examples) of OHIF. OHIF also provides thorough documentation with images and videos about how to use the image viewer and tools available.


```{r, fig.align='center', echo = FALSE, fig.alt= "The OHIF image viewer (center) includes a tool bar for to modify the viewing and annotation of images (above center), a menu on the left to access images related to the current patient in other studies, and a menu on the right for making measurements and exporting measurement reports.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd8d3f477a_159_2
")
```

For those interested, Gordon Harris and others are also working on a project called [Cornerstone](https://docs.cornerstonejs.org/), with the goal of providing software for others to display medical images in web browsers.


## PRISM

The Platform for Imaging in Precision Medicine called PRISM works behind the scenes in the Cancer Imaging Archive (TCIA) to allow users to work with the vast data available in TCIA, in terms of both imaging data and clinical information.  

According to Fred Prior:

> It is designed to collect, curate and manage Radiology and Pathology images, clinical information associated with those images, annotations and image derived feature sets.  PRISM is designed to run on a Kubernettes cluster, but all of the components are containerized so they can run stand-alone or in an alternate orchestration framework.


See this [article](https://ascopubs.org/doi/full/10.1200/CCI.20.00001) for more information.
