# Health care data sharing tools

<!-- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8178732/?report=classic -->
<!-- Amazon Comprehend -->
<!-- Healthcare Cost and Utilization Project (HCUP) -->

In this section we will cover the following learning objectives:

```{r, fig.alt = "Learning objectives: Explain how REDCap can be used to help keep data secure, compliant with HIPAA  requirements, and allows for collaboration; Identify best practices for keeping data safe using REDCap", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/10nOR2t1-F0E01fItN_l8uYRWslH2PmebPvhQzCBeCPM/edit?slide=id.g3896feb580f_34_14#slide=id.g3896feb580f_34_14")
```


### REDCap (Research Electronic Data Capture)

[REDCap](https://www.project-redcap.org/) is a very widely used browser-based software application for managing surveys and databases. It is very often used for clinical data. In fact, it is so widely used that there is a [conference](https://i2b2transmart.org/redcapcon2022/) dedicated to it.

REDCap allows for multi-institutional work, as well as compliance with HIPAA, [21 CFR Part 11](https://www.ecfr.gov/current/title-21/chapter-I/subchapter-A/part-11) for data for the FDA, [FISMA](https://www.techtarget.com/searchsecurity/definition/Federal-Information-Security-Management-Act) for government data, HIPAA, and [GDPR](https://gdpr-info.eu/) for data for the European Union. It was developed by a team at Vanderbilt University in 2004. It is not open-source, however it is free to use for non-commercial research [@redcap_2022].

You can find out more about how to use REDCap at the [REDCap website](https://projectredcap.org/) which includes instructional [videos](https://projectredcap.org/resources/videos/) and other resources.

There are several things to keep in mind when using REDCap from an ethical standpoint.

1) Roles

REDCap allows for various roles to be established for users on a project. Thus access to certain data and tasks can be restricted to certain individuals. As described previously, it is a good idea to restrict access to the smallest number of individuals necessary.

You can modify these roles using the `User Rights` menu.

```{r, fig.align='center', echo = FALSE, fig.alt= "REDCap User Rights Menu", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g133b14b2804_28_17")
```

This will first show you who has what role on the project and their rights. You can click on an individual role to modify it.

```{r, fig.align='center', echo = FALSE, fig.alt= "Menu items within the User Rights menu on REDcap", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g133b14b2804_28_22")
```

These roles should be verified by your institutional review board (IRB) before beginning a study. Changes to roles should also be reviewed by your IRB.

2) Reports

Reports that are exported can be customized to only show data that should be shared with the individual that you plan to share with. Please see the section on [de-identification](https://hutchdatascience.org/Ethical_Data_Handling_for_Cancer_Research/data-security.html#de-identification) to better understand what data you might want to be restrictive about sharing. Again, the way you intend to share your data should be reviewed by your IRB before you begin your study.

For example, you might remove the dates from the following report:

```{r, fig.align='center', echo = FALSE, fig.alt= "REDCap reports can be modified to remove identifiable information.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g133b14b2804_28_10")
```


3) Auditing

REDCap keeps track of all data modifications, as well as data exports or report generations, in addition to keeping track of who performs those actions. This can be helpful for checking what has happened and when, in case anything happens that is unexpected or unintended. This is also great from a reproducibility or transparency standpoint - you have a record of any modifications to the data. This information can be obtained from the `logging` menu.

4) Keep instruments short

If your instruments are too long, this can result in accidentally sharing data that you don't intend to, simply because you have more data to sift through. This also makes it easier to generate reports only on specific data that you would like to share.

5) Data can be locked

You can protect your data from accidentally being modified by locking specific data. Furthermore, at later stages of the project the data can no longer be modified.


```{r, fig.align='center', echo = FALSE, fig.alt= "Later project steps in REDCap can lock all data from being modified.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g133b14b2804_28_28")
```


<!-- Scott Carey can help with this section maybe -->

<div class = "warning">
Keep in mind that your institution likely has their own guidelines for how to use REDCap should you decide to use it. Also remember to verify what you plan to do with your institutional review board (IRB) before you begin the study.
</div>
